Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals

### 2018-10-24 paper/paper.tex
Minor changes to PE section.

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817. The effective one body formalism was previously shown to produce waveforms that agree with numerical simulations of aligned-spin BNS systems during the inspiral. This model includes aligned-spin effects, the quadrupolar and octopolar adiabatic tides, and the effect of dynamical tides. We improve this model by adding the spin-induced quadrupole moment and a prescription for tapering the end of the waveform to qualitatively agree with numerical simulations of the merger. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. Unfortunately, the current implementation takes several minutes to hours to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression to efficiently sample the parameter space. The resulting surrogate has a maximum mismatch of \red{$\sim 5\times 10^{-4}$} for the Advanced LIGO detector, and requires less than \red{0.3~s} to evaluate for a waveform with a starting frequency of 10~Hz (\red{0.04~s} for a starting frequency of 30~Hz). With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

In order to estimate the hyperparameters, we use the above assumption (Eq.~\eqref{eq:gaussian}) that the joint distribution of 
the data $y_i$ is a multivariate Gaussian with the following distribution:
\begin{equation}
\ln p(y_i | \bx_i, {\bm \theta}) = -\frac{1}{2} y_i (K^{-1})_{ij} y_j - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
\end{equation}
This is the log-likelihood for $y_i$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
given $y_i$ using Bayes' theorem
\begin{equation}
p({\bm \theta} | \bx_i, y_i) \propto p({\bm \theta}) p(y_i | \bx_i, {\bm \theta}).
\end{equation}
The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$. One can sample this posterior if interested in the distribution of hyperparameters $\btheta$. However, for the problem here, we simply want the maximum posterior. We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package~\cite{scikit-learn}. With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}. We note that once the vector $(K^{-1})_{ij}y_j$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
\caption{Projection of the training-set parameters onto the three-dimensional subspace $\{q, \chi_1, \chi_2\}$ (top) and the two-dimensional subspace $\{\xi_1, \xi_2\}$ (bottom). The 160 blue triangles were used for the initial training set (32 corners and 128 LHD points). The 400 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}) with the Mat\'{e}rn kernel. The 400 green circles were sampled uniformly from the parameter space.
}
\label{fig:LHD}
\end{figure}

@@

The surrogate evaluation time is the sum of two parts. The first part is the time needed to evaluate the residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ at each of the $N_A+N_\Phi$ interpolating nodes using GPR. Although optimizing the hyperparameters for GPR scales with the number of samples $N$ as $\mathcal{O}(N^3)$, the evaluation time of a stored GPR scales as $\mathcal{O}(N)$. The evaluation time for all the residuals therefore has a cost of $\mathcal{O}(N(N_A+N_\Phi))$. The second part is the time needed to resample the final surrogate (Eq.~\eqref{eq:amp_sur}--Eq.~\eqref{eq:hcross_sur}) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm low}$. \red{[[Michael: If you want, add details of how the specific implementation of resampling the waveform affects timing here.]]}
%\MP{Add cost of complex exponential sincos function: look at profiling results}
%\MP{Potentially, explore approximation to sincos for further speedup; need to guarantee that errors don't add up; range reduction is important for accuracy}

In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of starting frequency $f_{\rm low}$. The surrogate has a fixed time of \red{$\sim 50$~ms} to evaluate the GPR all of the nodes. \red{[[Details of how starting frequency depends on $f_{\rm low}$.]]} The additional cost to resample the amplitude and phase at uniformly-spaced frequencies with spline interpolation dominates at starting frequencies below $\sim 30$~Hz. We also compare the evaluation time to several BNS waveform models that have been used for LIGO data analysis. We note that at $\sim 30$~Hz, the \texttt{SEOBNRv4T\_surrogate} model is comparable to the \texttt{SEOBNRv4\_ROM\_NRTidal} and \texttt{IMRPhenomD\_NRTidal} models that were used in the analysis of the GW170817 event~\cite{BNSPE}. At lower frequencies, the surrogate is faster than all BNS waveform models except for \texttt{TaylorF2}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{timing.png}
\caption{Waveform evaluation time as a function of the waveform starting frequency. The time-domain waveforms
were sampled at 4096~Hz then Fourier transformed. The frequency-domain waveforms used the
same frequency samples as the Fourier transformed time-domain waveforms.}
\label{fig:timing}
\end{figure}

@@

Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now perform a parameter estimation run for GW170817. In the PE paper and catalog papers~\cite{}, the SEOBNRv4T surrogate was used with the parallelized PE code RapidPE. However, there were uncertainties about its ability to accurately recover the tidal parameter.

>>>>

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817. The effective one body formalism was previously shown to produce waveforms that agree with numerical simulations of aligned-spin BNS systems during the inspiral. This model includes aligned-spin effects, the quadrupolar and octopolar adiabatic tides, and the effect of dynamical tides. We improve this model by adding the spin-induced quadrupole moment and a prescription for tapering the end of the waveform to qualitatively agree with numerical simulations of the merger. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression to efficiently sample the parameter space. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

In order to estimate the hyperparameters, we use the above assumption (Eq.~\eqref{eq:gaussian}) that the joint distribution of 
the data $y_i$ is a multivariate Gaussian with the following distribution:
\begin{equation}
\ln p(y_i | \bx_i, {\bm \theta}) = -\frac{1}{2} y_i (K^{-1})_{ij} y_j - \frac{1}{2} \ln |K| - \frac{d}{2} \ln 2\pi.
\end{equation}
This is the log-likelihood for $y_i$ given the hyperparameters ${\bm \theta}$, and we can find the posterior for ${\bm \theta}$
given $y_i$ using Bayes' theorem
\begin{equation}
p({\bm \theta} | \bx_i, y_i) \propto p({\bm \theta}) p(y_i | \bx_i, {\bm \theta}).
\end{equation}
The prior $p({\bm \theta})$ is typically uniform and used to set the bounds on ${\bm \theta}$. One can sample this posterior if interested in the distribution of hyperparameters $\btheta$. However, for the problem here, we simply want the maximum posterior. We do this using the \texttt{gaussian\_process} module in the \texttt{scikit-learn} package~\cite{scikit-learn}. With these optimized hyperparameters, the final interpolating function is given by Eq.~\eqref{eq:mean} and its uncertainty by Eq.~\eqref{eq:var}. 
%We note that once the vector $(K^{-1})_{ij}y_j$ is precomputed, evaluating Eq.~\eqref{eq:mean} is a fast $\mathcal{O}(N)$ operation.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{trainingset3d.pdf}\\
\includegraphics[width=0.49\textwidth]{trainingset2d.pdf}
\caption{Projection of the training-set parameters onto the three-dimensional subspace $\{q, \chi_1, \chi_2\}$ (top) and the two-dimensional subspace $\{\xi_1, \xi_2\}$ (bottom). The 160 blue triangles were used for the initial training set (32 corners and 128 LHD points). The 400 red squares were generated using uncertainty sampling from the GPR error estimate (Eq.~\eqref{eq:errorrms}) with the Mat\'{e}rn kernel. The 400 green circles were sampled uniformly from the parameter space.}
\label{fig:LHD}
\end{figure}

@@

The surrogate evaluation time is dominated by two parts. The first part is the time needed to evaluate the residuals $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ and $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ at each of the $N_A+N_\Phi$ interpolating nodes using GPR. Although optimizing the hyperparameters for GPR scales with the number of training set samples $N$ as $\mathcal{O}(N^3)$, the evaluation time of a stored GPR scales as $\mathcal{O}(N)$ (see Eq.~\eqref{eq:mean}). The evaluation time for all the residuals therefore has a cost of $\mathcal{O}(N(N_A+N_\Phi))$. The second part is the time needed to resample the final surrogate (Eqs.~\eqref{eq:amp_sur}--~\eqref{eq:hcross_sur}) at uniformly spaced frequency samples in physical units beginning at a starting frequency $f_{\rm min}$.

In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for a equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor. The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz. At large starting frequencies the evaluation time is limited by the \red{$\sim 10$~ms} needed to calculate the GPR fits at each frequency node. As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed. The evaluation time is then dominated by the spline interpolation needed to resample the waveform. We also compare the evaluation time to three BNS waveform models used recently in the analysis of GW170817, (\texttt{TaylorF2}, \texttt{SEOBNRv4\_ROM\_NRTidal}, \texttt{IMRPhenomD\_NRTidal})~\cite{BNSPE} as well as the BBH model \texttt{SEOBNRv4\_ROM}. Only the analytic \texttt{TaylorF2} model is noticeably faster. In the bottom panel, we compare the evaluation time to the original time-domain \texttt{SEOBNRv4T} waveform model and find it is $\sim 1500$--6000 times faster.

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{../src/TEOB-LAL-implementation/timings/Models_timings_FD_monotonic.pdf}
\includegraphics[width=0.49\textwidth]{../src/TEOB-LAL-implementation/timings/SEOBNRv4T_surrogate_speedup_deltaF_monontonic.pdf}
\caption{Waveform evaluation time as a function of the waveform starting frequency $f_{\rm min}$. The time-domain \texttt{SEOBNRv4T} waveform was sampled at 4096~Hz then Fourier transformed. The frequency-domain waveforms used the same frequency samples as the Fourier transformed time-domain waveforms. Top panel: Evaluation times of the frequency-domain waveforms. Bottom panel: Speedup of \texttt{SEOBNRv4T\_surrogate} relative to the original \texttt{SEOBNRv4T}.}
\label{fig:timing}
\end{figure}

@@

Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817. In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE. However, there were uncertainties about its ability to accurately recover the tidal parameter.
